\vspace{-\baselineskip} Fast and scalable Gaussian process modeling with applications to astronomical time series \vspace{-2\baselineskip}

### 2017-06-11 paper/ms.tex
Minor edits; added discussion of power spectrum figure

To scale GP models to larger datasets, \citet{Rybicki:1995} presented a method
of computing the first term in \eq{gp-likelihood} in $\mathcal{O}(N)$
operations when the covariance function is given by
\begin{eqnarray}\eqlabel{kernel-simple}
k_\bvec{\alpha}(\tau_{nm}) = \sigma_n^2\,\delta_{nm} + a\,\exp(-c\,\tau_{nm})
\end{eqnarray}
where $\{{\sigma_n}^2\}_{n=1}^N$ are the measurement uncertainties,
$\delta_{nm}$ is the Kronecker delta, and $\bvec{\alpha} = (a,\,c)$.
The intuition behind this method is that, for this choice of $k_\bvec{\alpha}$,
the inverse of $K_\bvec{\alpha}$ is tridiagonal and can be computed
with a small number of operations for each data point.
Subsequently, \citet{Ambikasaran:2015} generalized this method to arbitrary
mixtures of exponentials
\begin{eqnarray}
k_\bvec{\alpha}(\tau_{nm}) = \sigma_n^2\,\delta_{nm} +
    \sum_{j=1}^J a_j\,\exp(-c_j\,\tau_{nm})
\end{eqnarray}
\response{by recognizing that the matrix $K_\bvec{\alpha}$ generated by this
kernel is semiseparable.}
In this case, the inverse is dense but \eq{gp-likelihood} can still be
evaluated with a scaling of $\mathcal{O}(N\,J^2)$, where $J$ is the number of
components in the mixture.
\response{In \sect{factor}, we derive a novel method with the same
$\mathcal{O}(N\,J^2)$ scaling that can be used when
$k_\bvec{\alpha}(\tau_{nm})$ is positive semi-definite.}

@@

The relationship between the \celeriteterm\ model and semiseparable linear
algebra was first recognized by Ambikasaran \citep{Ambikasaran:2015} building
on earlier work by Rybicki \& Press \citep{Rybicki:1995}.
Ambikasaran derived a scalable direct solver for any general semiseparable
matrix and applied this method to \celeriteterm\ models with $b_j = 0$ and
$d_j = 0$.
During the preparation of this paper, we generalized this earlier method to
include non-zero $b_j$ and $d_j$, but it turns out that we can derive a higher
performance algorithm by restricting our method to positive definite matrices.
In \app{psd}, we discuss methods for ensuring that a \celeriteterm\ model is
positive.

@@

The \celeriteterm\ model described in \sect{celerite} can be exactly
represented as a rank $R = 2\,J$ semiseparable matrix\footnote{We note that if
some \celeriteterm\ terms are real ($b_j=0$ and $d_j=0$) then these terms only
add one to the rank.
Therefore, if a \celeriteterm\ model has $J_\mathrm{real}$ real and
$J_\mathrm{complex}$ general terms, the semiseparable rank is only
$J_\mathrm{real} + 2\,J_\mathrm{complex}$.}
where the components of the relevant matrices are
\begin{eqnarray}\eqlabel{system}
U_{n,2\,j-1} &=& a_j\,e^{-c_j\,t_n}\,\cos(d_j\,t_n) +
    b_j\,e^{-c_j\,t_n}\,\sin(d_j\,t_n) \nonumber\\
U_{n,2\,j} &=& a_j\,e^{-c_j\,t_n}\,\sin(d_j\,t_n) -
    b_j\,e^{-c_j\,t_n}\,\cos(d_j\,t_n) \nonumber\\
V_{m,2\,j-1} &=& e^{c_j\,t_m}\,\cos(d_j\,t_m) \nonumber\\
V_{m,2\,j} &=& e^{c_j\,t_m}\,\sin(d_j\,t_m) \nonumber\\
A_{n,n} &=& \sigma_n^2 + \sum_{j=1}^J a_j \quad.
\end{eqnarray}
A na\"ive implementation of the algorithm in \eq{algorithm} for this system
will result in numerical overflow and underflow issues in many realistic cases
because both $U$ and $V$ have factors of the form $e^{\pm c_j t}$, where $t$
can be any arbitrarily large real number.
This issue can be avoided by reparameterizing the semiseparable representation
from \eq{system}.
%A similar issue was also noted by \citet{Ambikasaran:2015} and it can be
%avoided using a reparameterization.
We define the following pre-conditioned variables
\begin{eqnarray}\eqlabel{reparam}
\tilde{U}_{n,2\,j-1} &=& a_j\,\cos(d_j\,t_n) + b_j\,\sin(d_j\,t_n) \nonumber\\
\tilde{U}_{n,2\,j} &=& a_j\,\sin(d_j\,t_n) - b_j\,\cos(d_j\,t_n) \nonumber\\
\tilde{V}_{m,2\,j-1} &=& \cos(d_j\,t_m) \nonumber\\
\tilde{V}_{m,2\,j} &=& \sin(d_j\,t_m) \nonumber\\
\tilde{W}_{n,2\,j-1} &=& e^{-c_j\,t_n}\,W_{n,2\,j-1} \nonumber\\
\tilde{W}_{n,2\,j} &=& e^{-c_j\,t_n}\,W_{n,2\,j}
\end{eqnarray}
and the new set of variables
\begin{eqnarray}
\phi_{n,2\,j-1} = \phi_{n,2\,j} = e^{-c_j\,(t_n - t_{n-1})}
\end{eqnarray}
with the constraint
\begin{eqnarray}
\phi_{1,2\,j-1} = \phi_{1,2\,j} = 0 \quad.
\end{eqnarray}
Using this parameterization, we find the following, numerically stable,
algorithm to compute the Cholesky factorization of a \celeriteterm\ model
\begin{eqnarray}
    S_{n,j,k} &=& \phi_{n,j}\,\phi_{n,k}\,\left[S_{n-1,j,k} +
    D_{n-1,n-1}\,\tilde{W}_{n-1,j}\,\tilde{W}_{n-1,k}\right] \nonumber\\
D_{n,n} &=& A_{n,n} -
    \sum_{j=1}^{2\,J}\sum_{k=1}^{2\,J} \tilde{U}_{n,j}\,S_{n,j,k}\,\tilde{U}_{n,k}
    \nonumber\\
\tilde{W}_{n,j} &=& \frac{1}{D_{n,n}}\left[ \tilde{V}_{n,j} -
    \sum_{k=1}^{2\,J}\tilde{U}_{n,k}\,S_{n,j,k} \right] \quad.
\end{eqnarray}
As before, a single iteration of this algorithm requires $\mathcal{O}(J^2)$
operations and it is repeated $N$ times for a total scaling of
$\mathcal{O}(N\,J^2)$.
Furthermore, $\tilde{V}_{n,j}$ and $A_{n,n}$ can be updated in place and used
to represent $\tilde{W}_{n,j}$ and $D_{n,n}$, giving a total memory footprint
of $(6\,J + 1)\,N + J\,(J-1)/2$ floating point numbers.

@@

First, we empirically confirm that the method solves systems of the form
$K\,\bvec{z} = \bvec{y}$ to high numerical accuracy.
The left panel of \Figure{error} shows the L-infinity norm for the residual
$\bvec{z} - K^{-1}\,\bvec{y}$ for a range of dataset sizes $N$ and numbers of
terms $J$.
For each pair of $N$ and $J$ we repeated the experiment 10 times with
different randomly generated datasets and \celeriteterm\ parameters, and
averaged across experiments to remove some sampling noise from this figure.
The numerical error introduced by our method increases weakly for larger
datasets $\sim N^{0.15}$ and roughly linearly with $J$.

@@

We continue by measuring the real-world computational efficiency of this
method as a function of $N$ and $J$.
This experiment was run on a 2013 MacBook Pro with a dual 2.6 GHz Intel Core
i5 processor, but we find similar performance on other platforms.
The \celerite\ implementation used for this test is the \project{Python}
interface with a \project{C++} back end, but there is little overhead
introduced by \project{Python} so we find similar performance using
implementations in \project{C++} and \project{Julia} \citep{Bezanzon:2012}.
\Figure{benchmark} shows the computational cost of evaluating
\eq{gp-likelihood} using \celerite\ as a function of $N$ and $J$.
The empirical scaling as a function of data size matches the theoretical
scaling of $\mathcal{O}(N)$ for all values of $N$ and the scaling with the
number of terms is the theoretical $\mathcal{O}(J^2)$ for $J \gtrsim 10$, with
some overhead for smaller models.
For comparison, the computational cost of the same computation using the Intel
Math Kernel Library, a highly optimized general dense linear algebra package,
is shown as a dashed line.
\celerite\ is faster than the reference implementation for all but the
smallest datasets with many terms.

@@

We then sample the joint posterior probability (\eqalt{joint-post}) using
\emcee\ \citep{Goodman:2010, Foreman-Mackey:2013}.
We initialize \exampleinwalkers~walkers by sampling from three-dimensional
isotropic Gaussian centered the MAP parameter vector and with a standard
deviation of $10^{-4}$ in each dimension.
We then run \exampleinburn~steps of burn-in and \exampleinsteps~steps of MCMC.
To assess convergence, we estimate the mean integrated autocorrelation time of
the chain across parameters \citep{Sokal:1989, Goodman:2010} and find that the
chain results in \exampleineff~effective samples.

@@

As a demonstration, we fit a \celeriteterm\ model with a kernel given
by \eq{rot-kernel} to a \kepler\ light curve for the star KIC~1430163.
This star has a published rotation period of $3.88 \pm 0.58\,\mathrm{days}$,
measured using traditional periodogram and autocorrelation function approaches
applied to \kepler\ data from Quarters 0--16 \citep{Mathur:2014}.

@@

We initialize \exampleivnwalkers~walkers by sampling from an isotropic
Gaussian centered on the MAP parameters (the full set of parameters and their
priors are listed in Table~\ref{tab:example-4-params}), run
\exampleivnburn~steps of burn-in, and run \exampleivnsteps~steps of MCMC using
\emcee.
We estimate the mean autocorrelation time for the chains of
$\ln\nu_\mathrm{max}$ and $\ln\Delta\nu$ and find \exampleivneff\ effective
samples from the marginalized posterior density.
\Figure{astero-corner} shows the marginalized density for $\nu_\mathrm{max}$
and $\Delta\nu$ compared to the results from the literature.
This result is consistent within the published error bars and the posterior
constraints are tighter than the published results.
All asteroseismic analyses are known to have substantial systematic and
method-dependent uncertainties \citep{Verner:2011} so further experiments
would be needed to fully assess the reliability of this specific model.

@@

\section{Summary and discussion}

@@

Simulation of a Gaussian process for large $N$ requires multiplying an
$N$-dimensional vector of Gaussian normal deviates, $\bvec{q}$, by the square root
of the covariance matrix:
\begin{equation}
\bvec{y} = K_\bvec{\alpha}^{1/2} \bvec{q},
\end{equation}
where $q_i \sim N(0,1)$.  Note that $K_\bvec{\alpha} = K_\bvec{\alpha}^{1/2} \left(K_\bvec{\alpha}^{1/2}\right)^T$, which
comparing with the Cholesky factorization gives $K_\bvec{\alpha}^{1/2} = L D^{1/2}$,
where $(D^{1/2})_{ii} = \sqrt{D_{ii}}$.  Also note that the first term of $\mathcal{Q} = -2 \mathcal{L}$
with $\mu_\bvec{\theta} =0$ is given by
\begin{eqnarray}
\mathcal{Q} &=&\bvec{y}^T K_\bvec{\alpha}^{-1} \bvec{y}\\
&=& \bvec{y}^T \left(K_\bvec{\alpha}^{-1/2}\right)^T K_\bvec{\alpha}^{-1/2} \bvec{y}\\
&=& \bvec{q}^T K_\bvec{\alpha}^{1/2} \left(K_\bvec{\alpha}^{-1/2}\right)^T K_\bvec{\alpha}^{-1/2} K_\bvec{\alpha}^{1/2} \bvec{q}\\
&=& \bvec{q}^T \bvec{q}.
\end{eqnarray}
The last quantity is simply the dot product of an $N$-dimensional vector of independent standard normal random variables with itself,
which is distributed according to the chi-squared distribution with $N$ degrees of freedom, i.e.\ $\mathcal{Q} \sim \chi_N^2$; for
diagonal $K_\bvec{\alpha}$, $\mathcal{Q}$ is the $\chi^2$ statistic.

@@

Multiplication of the full covariance matrix times a vector can be accomplished in $\mathcal{O}(JN)$
operations with a stable recursive algorithm using the semiseparable form:
\begin{equation}
\bvec{y} =  \left(A + \mathrm{tril}(U V^T) + \mathrm{triu}(V U^T)\right) \bvec{z}.
\end{equation}
The algorithm for multiplication is given by
\begin{eqnarray}
f^+_{n,j} &=& \phi_{n+1,j} \left(f^+_{n+1,j} + \tilde U_{n+1,j} z_{n+1}\right)\\
f^-_{n,j} &=& \phi_{n,j} \left(f^-_{n-1,j} + \tilde V_{n-1,j} z_{n-1}\right)\\
y_n &=& A_{n,n} z_n + \sum_{j=1}^J \left[ \tilde V_{n,j} f^+_{n,j} + \tilde U_{n,j} f^-_{n,j}\right],
\end{eqnarray}
where $f^+_{N,j} = 0$ and $f^-_{1,j} = 0$ for all $j=1,...,2J$.  The recursion requires
two sweeps, a sweep downward from $n=N-1$ to $n=1$ to compute $f^+_{n,j}$ and a sweep upwards
from $n=2$ to $n=N$ to compute $f^-_{n,j}$.

Multiplication of the full covariance matrix is useful, for example, for isolating
a particular component of a kernel, which amounts to a customized filter of a dataset.
For example, a white-noise filter is given by:
\begin{eqnarray}
\bar f &=& K(\bvec{t},\bvec{t}) \left[K(\bvec{t},\bvec{t})+\sigma^2I\right]^{-1} \bvec{y},\\
&=& K(\bvec{t},\bvec{t}) \bvec{z}
\end{eqnarray}
where $\bvec{y}$ is the vector of measurements and $\bvec{z}$ may be computed from the Cholesky
decomposition as described above.  The second step multiplication of the white-noise-free
kernel times $\bvec{z}$ may be accomplished in $\mathcal{O}(JN)$ steps with this multiplication algorithm.
Note that for this particular example it is more efficient to compute $\bar f = \bvec{y} -\sigma^2 I \bvec{z}$,
but in general the filtering kernel will not differ from the full kernel by a diagonal matrix.

>>>>

To scale GP models to larger datasets, \citet{Rybicki:1995} presented a method
of computing the first term in \eq{gp-likelihood} in $\mathcal{O}(N)$
operations when the covariance function is given by
\begin{eqnarray}\eqlabel{kernel-simple}
k_\bvec{\alpha}(\tau_{nm}) = \sigma_n^2\,\delta_{nm} + a\,\exp(-c\,\tau_{nm})
\end{eqnarray}
where $\{{\sigma_n}^2\}_{n=1}^N$ are the measurement uncertainties,
$\delta_{nm}$ is the Kronecker delta, and $\bvec{\alpha} = (a,\,c)$.
The intuition behind this method is that, for this choice of $k_\bvec{\alpha}$,
the inverse of $K_\bvec{\alpha}$ is tridiagonal and can be computed
with a small number of operations for each data point.
Subsequently, \citet{Ambikasaran:2015} generalized this method to arbitrary
mixtures of exponentials
\begin{eqnarray}
k_\bvec{\alpha}(\tau_{nm}) = \sigma_n^2\,\delta_{nm} +
    \sum_{j=1}^J a_j\,\exp(-c_j\,\tau_{nm})
\end{eqnarray}
\response{by recognizing that the matrix $K_\bvec{\alpha}$ generated by this
kernel is semiseparable.}
In this case, the inverse is dense but \eq{gp-likelihood} can still be
evaluated with a scaling of $\mathcal{O}(N\,J^2)$, where $J$ is the number of
components in the mixture.
\response{In \sect{factor}, we derive a novel method with the same
$\mathcal{O}(N\,J^2)$ scaling that can be used when
$k_\bvec{\alpha}(\tau_{nm})$ is positive definite.}

@@

The relationship between the \celeriteterm\ model and semiseparable linear
algebra was first recognized by Ambikasaran \citep{Ambikasaran:2015} building
on earlier work by Rybicki \& Press \citep{Rybicki:1995}.
Ambikasaran derived a scalable direct solver for any general semiseparable
matrix and applied this method to \celeriteterm\ models with $b_j = 0$ and
$d_j = 0$.
During the preparation of this paper, we generalized this earlier method to
include non-zero $b_j$ and $d_j$, but it turns out that we can derive a higher
performance algorithm by restricting our method to positive definite matrices.
In \app{psd}, we discuss methods for ensuring that a \celeriteterm\ model is
positive definite.

@@

The \celeriteterm\ model described in \sect{celerite} can be exactly
represented as a rank $R = 2\,J$ semiseparable matrix\footnote{We note that if
some \celeriteterm\ terms are real ($b_j=0$ and $d_j=0$) then these terms only
add one to the rank.
Therefore, if a \celeriteterm\ model has $J_\mathrm{real}$ real and
$J_\mathrm{complex}$ general terms, the semiseparable rank is only
$R=J_\mathrm{real} + 2\,J_\mathrm{complex}=2J-J_{real}$.}
where the components of the relevant matrices are
\begin{eqnarray}\eqlabel{system}
U_{n,2\,j-1} &=& a_j\,e^{-c_j\,t_n}\,\cos(d_j\,t_n) +
    b_j\,e^{-c_j\,t_n}\,\sin(d_j\,t_n) \nonumber\\
U_{n,2\,j} &=& a_j\,e^{-c_j\,t_n}\,\sin(d_j\,t_n) -
    b_j\,e^{-c_j\,t_n}\,\cos(d_j\,t_n) \nonumber\\
V_{m,2\,j-1} &=& e^{c_j\,t_m}\,\cos(d_j\,t_m) \nonumber\\
V_{m,2\,j} &=& e^{c_j\,t_m}\,\sin(d_j\,t_m) \nonumber\\
A_{n,n} &=& \sigma_n^2 + \sum_{j=1}^J a_j \quad.
\end{eqnarray}
A na\"ive implementation of the algorithm in \eq{algorithm} for this system
will result in numerical overflow and underflow issues in many realistic cases
because both $U$ and $V$ have factors of the form $e^{\pm c_j t}$, where $t$
can be any arbitrarily large real number.
This issue can be avoided by reparameterizing the semiseparable representation
from \eq{system}.
%A similar issue was also noted by \citet{Ambikasaran:2015} and it can be
%avoided using a reparameterization.
We define the following pre-conditioned variables
\begin{eqnarray}\eqlabel{reparam}
\tilde{U}_{n,2\,j-1} &=& a_j\,\cos(d_j\,t_n) + b_j\,\sin(d_j\,t_n) \nonumber\\
\tilde{U}_{n,2\,j} &=& a_j\,\sin(d_j\,t_n) - b_j\,\cos(d_j\,t_n) \nonumber\\
\tilde{V}_{m,2\,j-1} &=& \cos(d_j\,t_m) \nonumber\\
\tilde{V}_{m,2\,j} &=& \sin(d_j\,t_m) \nonumber\\
\tilde{W}_{n,2\,j-1} &=& e^{-c_j\,t_n}\,W_{n,2\,j-1} \nonumber\\
\tilde{W}_{n,2\,j} &=& e^{-c_j\,t_n}\,W_{n,2\,j}
\end{eqnarray}
and the new set of variables
\begin{eqnarray}
\phi_{n,2\,j-1} = \phi_{n,2\,j} = e^{-c_j\,(t_n - t_{n-1})}
\end{eqnarray}
with the constraint
\begin{eqnarray}
\phi_{1,2\,j-1} = \phi_{1,2\,j} = 0 \quad.
\end{eqnarray}
Using this parameterization, we find the following, numerically stable,
algorithm to compute the Cholesky factorization of a \celeriteterm\ model
\begin{eqnarray}
    S_{n,j,k} &=& \phi_{n,j}\,\phi_{n,k}\,\left[S_{n-1,j,k} +
    D_{n-1,n-1}\,\tilde{W}_{n-1,j}\,\tilde{W}_{n-1,k}\right] \nonumber\\
D_{n,n} &=& A_{n,n} -
    \sum_{j=1}^{2\,J}\sum_{k=1}^{2\,J} \tilde{U}_{n,j}\,S_{n,j,k}\,\tilde{U}_{n,k}
    \nonumber\\
\tilde{W}_{n,j} &=& \frac{1}{D_{n,n}}\left[ \tilde{V}_{n,j} -
    \sum_{k=1}^{2\,J}\tilde{U}_{n,k}\,S_{n,j,k} \right] \quad.
\end{eqnarray}
As before, a single iteration of this algorithm requires $\mathcal{O}(J^2)$
operations and it is repeated $N$ times for a total scaling of
$\mathcal{O}(N\,J^2)$.
Furthermore, $\tilde{V}_{n,j}$ and $A_{n,n}$ can be updated in place and used
to represent $\tilde{W}_{n,j}$ and $D_{n,n}$, giving a total memory footprint
of $(6\,J + 1)\,N + J\,(J-1)/2$ floating point numbers.

@@

First, we empirically confirm that the method solves systems of the form
$K\,\bvec{z} = \bvec{y}$ to high numerical accuracy.
The left panel of \Figure{error} shows the L-infinity norm for the residual
$\bvec{z} - K^{-1}\,\bvec{y} = \bvec{z}-K^{-1}K\bvec{z}$ for a range of dataset sizes $N$ and numbers of terms $J$.
For each pair of $N$ and $J$ we repeated the experiment 10 times with
different randomly generated datasets and \celeriteterm\ parameters, and
averaged across experiments to remove some sampling noise from this figure.
The numerical error introduced by our method increases weakly for larger
datasets $\sim N^{0.15}$ and roughly linearly with $J$.

@@

We continue by measuring the real-world computational efficiency of this
method as a function of $N$ and $J$.
This experiment was run on a 2013 MacBook Pro with a dual 2.6 GHz Intel Core
i5 processor, but we find similar performance on other platforms.
The \celerite\ implementation used for this test is the \project{Python}
interface with a \project{C++} back end, but there is little overhead
introduced by \project{Python} so we find similar performance using
the implementation in \project{C++} and an independent implementation in
\project{Julia} \citep{Bezanzon:2012}.
\Figure{benchmark} shows the computational cost of evaluating
\eq{gp-likelihood} using \celerite\ as a function of $N$ and $J$.
The empirical scaling as a function of data size matches the theoretical
scaling of $\mathcal{O}(N)$ for all values of $N$ and the scaling with the
number of terms is the theoretical $\mathcal{O}(J^2)$ for $J \gtrsim 10$, with
some overhead for smaller models.
For comparison, the computational cost of the same computation using the Intel
Math Kernel Library, a highly optimized general dense linear algebra package,
is shown as a dashed line, which is independent of $J$.
\celerite\ is faster than the reference implementation for all but the
smallest datasets with many terms, $J \ga 64$ for $N\approx 512$, 
and $J$ increasing with larger $N$.

@@

We then sample the joint posterior probability (\eqalt{joint-post}) using
\emcee\ \citep{Goodman:2010, Foreman-Mackey:2013}.
We initialize \exampleinwalkers~walkers by sampling from a three-dimensional
isotropic Gaussian centered the MAP parameter vector and with a standard
deviation of $10^{-4}$ in each dimension.
We then run \exampleinburn~steps of burn-in and \exampleinsteps~steps of MCMC.
To assess convergence, we estimate the mean integrated autocorrelation time of
the chain across parameters \citep{Sokal:1989, Goodman:2010} and find that the
chain results in \exampleineff~effective samples.

@@

As a demonstration, we fit a \celeriteterm\ model with a kernel given
by \eq{rot-kernel} to a \kepler\ light curve for the star KIC~1430163.
This star has a published rotation period of $3.88 \pm 0.58\,\mathrm{days}$,
measured using traditional periodogram and autocorrelation function approaches
applied to \kepler\ data from Quarters 0--16 \citep{Mathur:2014}, covering
about four years.

@@

We initialize \exampleivnwalkers~walkers by sampling from an isotropic
Gaussian centered on the MAP parameters (the full set of parameters and their
priors are listed in Table~\ref{tab:example-4-params}), run
\exampleivnburn~steps of burn-in, and run \exampleivnsteps~steps of MCMC using
\emcee.
We estimate the mean autocorrelation time for the chains of
$\ln\nu_\mathrm{max}$ and $\ln\Delta\nu$ and find \exampleivneff\ effective
samples from the marginalized posterior density.
\Figure{astero-corner} shows the marginalized density for $\nu_\mathrm{max}$
and $\Delta\nu$ compared to the results from the literature.
This result is consistent within the published error bars and the posterior
constraints are tighter than the published results.
\Figure{astero_psd} shows a comparison of
the power spectrum estimated with a Lomb-Scargle periodogram 
with the power spectrum inferred with \celerite.  Despite the much smaller 
one-month sequence of data used in the \celerite\ analysis, the power spectrum 
inferred with \celerite\ matches well the periodogram estimated from
the full four years of Kepler data.
All asteroseismic analyses are known to have substantial systematic and
method-dependent uncertainties \citep{Verner:2011} so further experiments
would be needed to fully assess the reliability of this specific model.

@@

\section{Discussion}

@@

Simulation of a Gaussian process for large $N$ requires multiplying an
$N$-dimensional vector of Gaussian normal deviates, $\bvec{q}$, by the square root
of the covariance matrix:
\begin{equation}
\bvec{y} = K_\bvec{\alpha}^{1/2} \bvec{q},
\end{equation}
where $q_i \sim N(0,1)$.  Note that $K_\bvec{\alpha} = K_\bvec{\alpha}^{1/2} \left(K_\bvec{\alpha}^{1/2}\right)^T$, which
comparing with the Cholesky factorization gives $K_\bvec{\alpha}^{1/2} = L D^{1/2}$,
where $(D^{1/2})_{ii} = \sqrt{D_{ii}}$.  Also note that the first term of $\mathcal{Q} = -2 \mathcal{L}$
with $\mu_\bvec{\theta} =0$ is given by
\begin{eqnarray}
\mathcal{Q} &=&\bvec{y}^T K_\bvec{\alpha}^{-1} \bvec{y}\\
&=& \bvec{y}^T \left(K_\bvec{\alpha}^{-1/2}\right)^T K_\bvec{\alpha}^{-1/2} \bvec{y}\\
&=& \bvec{q}^T K_\bvec{\alpha}^{1/2} \left(K_\bvec{\alpha}^{-1/2}\right)^T K_\bvec{\alpha}^{-1/2} K_\bvec{\alpha}^{1/2} \bvec{q}\\
&=& \bvec{q}^T \bvec{q}.
\end{eqnarray}
The last quantity is simply the dot product of an $N$-dimensional vector of independent standard normal random variables with itself,
which is distributed according to the chi-squared distribution with $N$ degrees of freedom, i.e.\ $\mathcal{Q} \sim \chi_N^2$; for
diagonal $K_\bvec{\alpha}$, $\mathcal{Q}$ becomes the $\chi^2$ statistic.

@@

Multiplication of the full covariance matrix times a vector can be accomplished in $\mathcal{O}(RN)$
operations with a stable recursive algorithm using the semiseparable form:
\begin{equation}
\bvec{y} =  \left(A + \mathrm{tril}(U V^T) + \mathrm{triu}(V U^T)\right) \bvec{z}.
\end{equation}
The algorithm for multiplication is given by
\begin{eqnarray}
f^+_{n,j} &=& \phi_{n+1,j} \left(f^+_{n+1,j} + \tilde U_{n+1,j} z_{n+1}\right)\\
f^-_{n,j} &=& \phi_{n,j} \left(f^-_{n-1,j} + \tilde V_{n-1,j} z_{n-1}\right)\\
y_n &=& A_{n,n} z_n + \sum_{j=1}^R \left[ \tilde V_{n,j} f^+_{n,j} + \tilde U_{n,j} f^-_{n,j}\right],
\end{eqnarray}
where $f^+_{N,j} = 0$ and $f^-_{1,j} = 0$ for all $j=1,...,R$.  The recursion requires
two sweeps, a sweep downward from $n=N-1$ to $n=1$ to compute $f^+_{n,j}$ and a sweep upwards
from $n=2$ to $n=N$ to compute $f^-_{n,j}$.

Multiplication of the full covariance matrix is useful, for example, for isolating
a particular component of a kernel, which amounts to a customized filter of a dataset.
For example, a white-noise filter is given by:
\begin{eqnarray}
\bar f &=& K(\bvec{t},\bvec{t}) \left[K(\bvec{t},\bvec{t})+\sigma^2I\right]^{-1} \bvec{y},\\
&=& K(\bvec{t},\bvec{t}) \bvec{z}
\end{eqnarray}
where $\bvec{y}$ is the vector of measurements and $\bvec{z}$ may be computed from the Cholesky
decomposition as described above.  The second step multiplication of the white-noise-free
kernel times $\bvec{z}$ may be accomplished in $\mathcal{O}(RN)$ steps with this multiplication algorithm.
Note that for this particular example it is more efficient to compute $\bar f = \bvec{y} -\sigma^2 I \bvec{z}$,
but in general the filtering kernel will not differ from the full kernel by a diagonal matrix.
